---
title: "KwaThunzi runoff time-series simulation"
output: rmarkdown::github_document
---

```{r libraries, include=FALSE}
library(readr)
library(lubridate)
library(ggplot2)
library(hydrostats)
```

## Data import and preparation
Import the runoff dataset 1960-2018 of the Umkomazi gauging station (ID: U1H005) available at the [Department of Water Affairs](http://www.dwa.gov.za/Hydrology/). The runoff Q is in unit [m³/s].
```{r load data, message=FALSE, warning=FALSE}
runoff <- read_table2("../data/U1H005_daily.txt", 
                      col_types = cols(DATE = col_integer()), 
                      skip = 10)
head(runoff, 3)
```

Clean the data and set column formats. The quality code indicating the validity of each observation is available [here]()
Create seperate columns for 'year' and 'doy'(day-of-year) for each observation.
```{r data cleaning}                      
runoff[is.na(runoff$QUAL),2] <- NA # remove Quality measures that shifted in the Q column
runoff <- ts.format(runoff, format="%Y%m%d", cols=c(1,2)) # set date format
runoff$year <- as.integer(substr(runoff$Date, 1,4)) # derive year
runoff$doy <- as.numeric(strftime(runoff$Date, format = "%j"))  # derive day-of-year
runoff <- na.omit(runoff) # delete empty fields
head(runoff, 3)
```
```{r plot time series, echo=FALSE}
ggplot(runoff, aes(x=Date, y=Q)) + 
  ylim(0, 400) +
  ylab("Q [m³/s]") +
  geom_line(aes(group=1), colour="blue", size=.5) +
  facet_wrap(runoff$year, ncol=10, scales="free_x") +
  scale_x_datetime(date_breaks = "1 month", minor_breaks=NULL) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 6))
```

## Runoff transformation and aggregation

Transform discharge by base 10 logarithm.
```{r log tranformation}
runoff$Qlog <- log10(runoff$Q)  # calculate Log base 10 of runoff
runoff <- runoff[!(runoff$Qlog=='-Inf'),] # delete -Inf
head(runoff,3)
```
```{r log tranformation plot, echo=FALSE}
# Density of logarithmic runoff (per doy)  
ggplot(runoff, aes(x=doy,y=Qlog)) +
  geom_bin2d(binwidth=c(10,0.1)) +
  labs(x="Day of year", y="Log Q [m³/s]", title="Observed runoff Mkomazi (U1H005)") +
  ylim(-2,3)
  
```


Calculate daily mean and standard deviation for each day-of-year. 
```{r aggregate runoff, warning=FALSE}
#Create empty data frame
runoff_aggregates <- data.frame()

#aggregate mean runoff per day
runoff_aggregates <- aggregate(runoff, by=list(runoff$doy), FUN=mean)[,c('doy','Qlog')]

#aggregate mean runoff per day
runoff_aggregates$sd <- aggregate(runoff, list(runoff$doy), FUN=sd)[,c('Qlog')]

names(runoff_aggregates) <- c('doy','mean','sd')

```

## Runoff simulation
Simulate expected runoff of each day of any timeseries. Create randomized values, based on the observed annual discharge distribution, which was derived from the time series 1960-2018. The random values are estimated from mean and standard deviation of each daily runoff.
```{r simulation, message=FALSE, warning=FALSE}
set.seed(2020)
number_of_years <- 500 # years to be simulated
runoff_pred <- data.frame()

for(i in runoff_aggregates$doy){
  x <- rnorm(number_of_years, mean=runoff_aggregates$mean[i], sd=runoff_aggregates$sd[i])
  pred <- data.frame(cbind(seq(1,500,1), # year
                         rep(i,500),   # doy
                         x))           # predicted values
  runoff_pred <- rbind(runoff_pred, pred)
}
names(runoff_pred) <- c('year','doy','Qlog')
```
```{r simulation plot, echo=FALSE}
ggplot(runoff_pred, aes(x=doy,y=Qlog)) +
  geom_bin2d(binwidth=c(10,0.1)) +
  labs(x="Day of year", y="Log Q [m³/s]", title="Simulated runoff Mkomazi (U1H005)")+
  ylim(-2,3)
```
Backtransformation of the logarithmic runoffs to regular Q [m³/s]
```{r backtransformation, warning=FALSE}
runoff_pred$Q <- 10^runoff_pred$Qlog
```

## Calculation of area specific runoff and export to file
Calculate area specific runoff (q) after Baumgartner and Liebscher (1996) and Casper and Bormann (2016) using
\begin{equation}q = \frac{Q}{A}\end{equation} with Q = runoff at the gauging station and A = catchment area
```{r area specific runoff}
catchment_area <- 1744 * 1000000 # area in [m²]
runoff_pred$q <- runoff_pred$Q/catchment_area
```

Order rows by year and day-of-year
```{r order}
runoff_pred <- runoff_pred[order(runoff_pred$year, runoff_pred$doy),]
head(runoff_pred, 6)
```
Export data to a structured file with the two columns 'year' and 'q', which is ready to use with the gully model
```{r export}
out <- runoff_pred[,c("year", "q")] 
out$q <- format(out$q, scientific = FALSE) # change to decimals
write.table(out, "../output/S_150_lts.txt", sep="\t", row.names=F, col.names=F, quote=F)
```
## References
BAUMGARTNER, A. & LIEBSCHER, H. J. 1996. Lehrbuch der Hydrologie, Band 1: Allgemeine Hydologie, Quantitative Hydrologie, Stuttgart, Germany, Schweizerbarth.
CASPER, M. & BORMANN, H. 2016. Abfluss im Gewässersystem. In: FÖHRER, N., BORMANN, H., MIEGEL, K., CASPER, M., BRONSTERT, A., SCHUMANN, A. & WEILER, M. (eds.) Hydrologie. Bern, Switzerland: Haupt.
