---
title: "KwaThunzi runoff time-series simulation"
author: "Christian Sommer"
output:
  md_document:
    variant: markdown_github
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_float: false
---
# Content
This is a an R script to simulate a runoff time-series based on empirical data for the DIMGUL gully model by Alexey Sidorchuk (1998, 1999). In a first step, observed flow data are transformed and a temporaly variable gaussian function is derived. In a second step, random values are derived from this function in a reproducible manner, in order to simulate a long-term time-series, which resembles the natural conditions, whille providing data, where no observations are available.
The simulated time-series reflects the conditions of the observed data, *without effects like climate change or land use / land cover change, etc*. In the third step, we calculate hydrological values like area specific runoff q and runoff height R. In the last step, data are exportet in a structered text file, which is ready-to-use with the DIMGUL implementation.

## Units

| Name                                 | Abbr | Unit                 |
|--------------------------------------|------|----------------------|
| Discharge                            | Q    | m³/s                 |
| logarithmic Discharge                | Qlog | log(m³/s)            |
| area specific runoff "Abflussspende" | q    | l/(m²s)              |
| daily runoff height "Abflusshöhe"    | R    | mm/day               |
| subcatchment area                    | A    | km²                  |

# Calculation
```{r libraries, include=FALSE}
library(readr)
library(lubridate)
library(ggplot2)
library(hydrostats)
```

## Data import and preparation
Import the runoff dataset 1960-2018 of the Umkomazi gauging station (ID: U1H005) available at the [South African Department of Water Affairs](http://www.dwa.gov.za/hydrology/Verified/HyDataSets.aspx?Station=U1H005). The runoff Q is in unit [m³/s].
```{r load data, message=FALSE, warning=FALSE}
runoff <- read_table2("../data/U1H005_daily.txt", 
                      col_types = cols(DATE = col_integer()), 
                      skip = 10)
```
```{r echo=FALSE}
head(runoff, 3)
```

We clean the data and set column formats. The quality code indicates the validity of each observation and is available [here](http://www.dwa.gov.za/hydrology/Verified/HyCodes.aspx).
Then, we derive seperate columns for 'year' and 'doy'(day-of-year) for each observation with the package [Lubridate](https://lubridate.tidyverse.org)
```{r data cleaning}                      
runoff[is.na(runoff$QUAL),2] <- NA # remove Quality measures that shifted in the Q column
runoff <- ts.format(runoff, format="%Y%m%d", cols=c(1,2)) # set date format
runoff$year <- as.integer(substr(runoff$Date, 1,4)) # derive year
runoff$doy <- as.numeric(strftime(runoff$Date, format = "%j"))  # derive day-of-year
runoff$month <- as.numeric(strftime(runoff$Date, format = "%m"))  # derive day-of-year
runoff <- na.omit(runoff) # delete empty fields
```

As a result, we get this table...
```{r echo=FALSE}
head(runoff, 3)
```

... and this continuous dataset of the runoff at the gauging station.
```{r plot time series, echo=FALSE}
ggplot(runoff, aes(x=Date, y=Q)) + 
  ylim(0, 400) +
  ylab("Q [m³/s]") +
  geom_line(aes(group=1), colour="blue", size=.5) +
  facet_wrap(runoff$year, ncol=10, scales="free_x") +
  scale_x_datetime(date_breaks = "1 month", minor_breaks=NULL) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 6))
```


## Calculation of area specific runoff and runoff height
Until here, we were dealing with runoff values of the gauging station. Now we relate the results to the entire subcatchment area. Therefore, we calculate the **area specific runoff (q)** after Baumgartner & Liebscher (1996: eq. 14.1, p. 475) and Casper & Bormann (2016: eq. 8.4, p. 128) using
\begin{equation}q = \frac{Q}{A}\end{equation} 
with Q = runoff at the gauging station [m³/s] and A = subcatchment area of the gauging station [km²]. The area specific runoff q is in [l/(km²s)].
```{r area specific runoff}
catchment_area <- 1744 # area in [km²]
runoff$q <- runoff$Q*1000/catchment_area # Q [m³/s] to [l/(km²/s)]
```

We also calculate the accumulated **runoff R** for each day in [mm/d] after Baumgartner & Liebscher (1996: eq. 14.2, p. 475)
\begin{equation}R = \frac{q⋅t}{10^6}\end{equation}
with t = time in [s]
```{r accumulated daily runoff, message=FALSE, warning=FALSE}
runoff$R <- runoff$q * 86400 / 10^6 # per day
```

Order rows by year and day-of-year
```{r order}
runoff <- runoff[order(runoff$year, runoff$doy),]
```

This is the resulting table with simulated runoff in the differen formats Q [m³/s], q [l/(km²s)] and R [mm/d]:
```{r echo=FALSE}
head(runoff)
```

# File Export
Finally, we export the simulated data to a structured file with the two columns 'year' and runoff ('q' or 'R'), which is ready to use with the gully model. Note that we convert q from unit [l/(km²s)] to [l/(m²s)] for compability reasons.
```{r export}
# Export q in [l/(km²s)]
out <- runoff[,c("year", "q")]
out$q <- out$q/1000000 # [l/(km²s)] to [l/(m²s)]
out$q <- format(out$q, scientific = FALSE) # change to decimals
write.table(out, "../output/S_150_obs_Lsqm.txt", sep="\t", row.names=F, col.names=F, quote=F)

# Export h in [mm/d]
out <- runoff[,c("year", "R")] 
out$R <- format(out$R, scientific = FALSE) # change to decimals
write.table(out, "../output/S_150_obs_lts_R_mmd.txt", sep="\t", row.names=F, col.names=F, quote=F)
```

# Useful Links
- [https://www.bauformeln.de/einheiten-rechner/niederschlagsintensitaet-und-abflussspende/](https://www.bauformeln.de/einheiten-rechner/niederschlagsintensitaet-und-abflussspende/)

# References
BAUMGARTNER, A. & LIEBSCHER, H. J. 1996. Lehrbuch der Hydrologie, Band 1: Allgemeine Hydologie, Quantitative Hydrologie, Stuttgart, Germany, Schweizerbarth.

CASPER, M. & BORMANN, H. 2016. Abfluss im Gewässersystem. In: FÖHRER, N., BORMANN, H., MIEGEL, K., CASPER, M., BRONSTERT, A., SCHUMANN, A. & WEILER, M. (eds.) Hydrologie. Bern, Switzerland: Haupt.

SIDORCHUK, A. & SIDORCHUK, A. (1998). Model for estimating gully morphology Modelling Soil Erosion, Sediment Transport and Closely Related Hydrological Processes, Vienna.

SIDORCHUK, A. (1999). "Dynamic and static models of gully erosion." CATENA 37(3): 401-414.