---
title: "Soil parameters"
author: "Christian Sommer"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_float: false
  md_document:
    variant: markdown_github
---

# Soil data
We start by importing the results of a soil texture analysis from a table. The table contains a row for each layer with the following columns:

- Name
- lab number
- thickness in [cm]
- texture in [%]
- organic matter in [%]
- comment

```{r}
d <- read.csv("../data/soil_profile.csv", sep=";")
```


```{r echo=FALSE}
head(d)
```
Here are the results plotted in a USDA soil texture triangle
```{r texture triangle, echo=FALSE, fig.height=6, message=FALSE, warning=FALSE}
library(soiltexture)
# Calculate Silt and Sand Fractions, all columns must be capitals
x <- d
x$SILT <- d$csilt + d$fsilt
x$SAND <- d$csand + d$msand + d$fsand + d$vfsand
x$CLAY <- d$clay
x <- TT.normalise.sum(tri.data=x)


geo <- TT.plot(tri.data=x,
              class.sys = "USDA.TT",
              #class.p.bg.col = TRUE,
              text.sum = 100,
              main="USDA",
              col="#A51E37",
              #arrows.show = F,
              pch=15,
)


TT.text(tri.data = x,
        geo=geo,
        #labels = c("L1", "L2", "L3", "L4"),
        labels = seq(1,nrow(d)),
        cex=1,
        font=2,
        col="#A51E37",
        #adj=1,
        offset=0.5,
        pos=2)
rm(x)
```


# Rill erodibility
Is calculated after Alberts et al. (1995) using the proportion of very fine sand $VFS$ [%], clay $CL$ [%] and the content of organic matter $OM$ [%]. If the soil contains more than 30% sand, we apply this equation: 
\begin{equation}K_{r} = 0.00197 + 0.00030 VFS + 0.038633 \mathrm{e}^{-1.84OM}\end{equation}
If the sand content is lower than 30%, the authors suggest using:
\begin{equation}K_{r} = 0.0069 + 0.134 \mathrm{e}^{-20CL}\end{equation}
```{r rill erodibility function, message=FALSE, warning=FALSE}
erodibility <- function(clay, vfsand, om, sand){
  if(sand >= 30){
    erodibility <- 0.00197+0.00030*vfsand+0.038633*exp(-1.84*om)
  }else if(sand < 30){
    erodibility <- 0.0069+0.134*exp(-20*clay)
  }
}
d$Kr <- erodibility(vfsand=d$vfsand, om=d$som, clay=d$clay, sand=sum(d$csand, d$msand, d$fsand, d$vfsand))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$Kr)
```


# Critical velocity
## Mean particle size diameter
We calculate the mean particle size diameter $Dg% in [m] using the equation by Renard et al. (1997) with the primary fraction of the grain size classes $f_{i}$ in [%] and the arithmetic mean of the borders of the grain size classes $m_{i}$ in [mm].
\begin{equation} Dg = exp(0.01 \sum{} f_{i}\cdot ln \cdot m_{i})\end{equation}
```{r mean_particle_size_diameter, message=FALSE, warning=FALSE}
mean_particle_diameter <- function(clay=0, silt=0, sand=0, csilt=0, fsilt=0, csand=0, msand=0, fsand=0, vfsand=0, method="geometric", classification="USDA"){
  if(classification=="USDA"){
    gsize <- c(mean(c(0, 0.002)),    # Clay
               mean(c(0.002, 0.05)), # Silt
               mean(c(0.05, 2)),   # Sand
               mean(c(0.05, 0.02)),  # Coarse silt
               mean(c(0.02, 0.002)), # Fine silt
               mean(c(2, 0.5)),      # Coarse sand
               mean(c(0.5, 0.25)),   # Medium sand
               mean(c(0.25, 0.1)),   # Fine sand
               mean(c(0.1, 0.05)))   # Very fine sand    
  }
  if(method=="geometric"){
    mean_diameter <- exp(0.01 * (
        clay * log(gsize[1]) +
        silt * log(gsize[2]) +
        sand * log(gsize[3]) +
        csilt * log(gsize[4]) +
        fsilt * log(gsize[5]) +
        csand * log(gsize[6]) +  
        msand * log(gsize[7]) +  
        fsand * log(gsize[8]) +  
        vfsand * log(gsize[9])
    ))
  }
  return(mean_diameter/1000) # from mm to m
}

d$Dg <- mean_particle_diameter(d$clay, d$csilt, d$fsilt, d$csand, d$msand, d$fsand, d$vfsand, method="geometric", classification="USDA")
```
Note, that the Mean particle diameter is converted from [mm] to [m] for further processing
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$Dg)
```


## Shear stress
There are multiple ways to estimate Shear stress $\tau_{c}$ from soil texture. However, you'll notice the large differences between the results of the different equations. We'll continue the method from Alberts et al. (1995).

### after Smerdon (1961)
\begin{equation} \tau_{c} = 3.11 \cdot 10^{0.0182C}\end{equation}
```{r shear stress smeardon, message=FALSE, warning=FALSE}
shear_stress <- function(clay){
  shear_stress <- 3.11*10^(0.0182*clay) # in Pa (Pascal)
}

d$tau <- shear_stress(d$clay)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$tau)
```

### after Alberts et al. (1995)
\begin{equation} \tau_{c} = 2.67 + 0.065C - 0.058VFS\end{equation}
```{r shear stress Alberts, message=FALSE, warning=FALSE}
shear_stress <- function(clay, vfsand){
  shear_stress <- 2.67 + 0.065 * clay - 0.058 * vfsand
}

d$tau <- shear_stress(d$clay, d$vfsand)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$tau)
```


## Critical velocity after Bogomolov & Mikhailov (1988)
Finally, we've collected all parameters to calculate the critical velocity of erosion initiation $V_{crit}$ [m/s] using the equation from  Bogomolov & Mikhailov (1988):
\begin{equation}V_{crit} = 1.25\sqrt{\frac{2m_{1}}{1.54pn_{1}}[(p_{s}-p)gd + 1.25C^{1}_{f}K_{0}]}\end{equation}
The equation includes some predefined parameter values, obtained from Märker (2001) and Sidrochuk (2019, pers. comm.), comprising the colloidal suspended load factor $m_{1}$, the turbulence parameter $n_{1}$, the density of water $p$, the gravity constant $g$ and the soil mechanical coefficent $K_{0}$.
The colloidal suspended load factor can be obtained from Mirtshkulava (1988):

| colloidal factor                             | m    | colloidal factor                             | m    |
|----------------------------------------------|------|----------------------------------------------|------|
| colloidal particle concentration > 0.1 kg/m³ | 1.4  | humid climates                               | 0.71 |
| coarse bedload                               | 0.8  | light meandering channels                    | 0.95 |
| water plants                                 | 1.15 | medium meandering channels                   | 0.85 |
| arid climates                                | 0.22 | strong meandering channels                   | 0.65 |

```{r, critical velocity, message=FALSE, warning=FALSE}
v_crit <- function(m1=1,     # 1=clear water,...,1.4=colloidal suspended load > 0.1 kg/m3
                   n1=4,     # Turbulence: 4
                   ps,       # Density of soil kg/m3
                   p=1000,   # Density of water: ~1000 kg/m3
                   g=9.81,   # Newton/kg
                   d,        # mean particle size diameter (m)
                   K0=1,     # Soil mechanical coefficent: 0.5 (Maerker 2002) or 1 (Sidorchuk 2019)
                   Cnf       # Soil shear strength (Pa)
){
  # Critical velocity of bed erosion initiation (after 1pers. comm. Sidorchuck 2019)
  v_crit <- 1.25 * sqrt(
    (2*m1/1.54*p*n1)*
      ((ps-p)*g*d + 1.25*Cnf*K0)
  )
  return(v_crit)
}

d$v_crit <- v_crit(ps=1200, d=d$Dg, Cnf=d$tau)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$v_crit)
```

We can compare our results with the following results from Bogomolov & Mikhaylov (1972, after Sidorchuk 1999).

| Soil texture      | Vcr  | Soil texture      | Vcr  | Soil texture      | Vcr  | Soil texture      | Vcr  |
|-------------------|------|-------------------|------|-------------------|------|-------------------|------|
| Large   boulders  | 4    | Very coarse sand  | 0.65 | Loamy sand        | 0.6  | Hard loam         | 1.15 |
| Small boulders    | 3.2  | Coarse sand       | 0.6  | Soft sandy loam   | 0.7  | Soft clay         | 0.8  |
| Coarse gravel     | 1.1  | Medium sand       | 0.57 | Hard sandy loam   | 1    | Medium hard clay  | 1.2  |
| Medium gravel     | 0.9  | Fine sand         | 0.32 | Soft loam         | 0.75 | Hard clay         | 1.4  |
| Fine gravel       | 0.75 | Silt              | 0.55 | Medium hard loam  | 1    | Very hard clay    | 1.7  |


# Critical slope
The critical slope is simply the slope measured in the field or obtained from a high-res DEM coverted to radians.
```{r critical slope, message=FALSE, warning=FALSE}
s_crit <- function(slope, unit="degrees"){
  if(unit=="degrees"){
    s_crit <- slope*pi/180
  }else if(unit=="percent"){
    s_crit <- atan(slope/100)
  }
}
# Since all layers have the same slope in this example, we can assign the same value. Otherwise get the value from the table.
d$s_crit <- s_crit(57, "degrees")
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(d$s_crit)
```



# Export
Now we can export all relevant variables of each layer to an individual PARAM.txt parameter file, ready for DYMGUL.
```{r export, message=FALSE, warning=FALSE}
layer_export <- function(years, lnumber=1, v_crit, s_crit, Kr, Qdat="S_150.txt"){
  for (i in seq(1,lnumber)){
    filename <- paste("../output/PARAM", toString(i), ".txt", sep="")
    Layer <- file(filename, open="w")
    lines <- c("0\t0", 
               i,              # Number of layer
               v_crit[i],      # Critical Velocity
               "0",            # Index of texture layer (unused)
               s_crit[i],      # Stable slope
               Kr[i],          # Erodibility coefficent
               "S_150.txt",
               years,
               seq(1, years, 1))
    writeLines(lines, con=Layer)
    close(Layer)
    rm(i, filename, Layer, lines)
}}


# Apply
layer_export(years=59, lnumber=nrow(d), v_crit=d$v_crit, s_crit=d$s_crit, Kr=d$Kr)
```




# References
Alberts E. E., Nearing M. A., Weltz M. A., Risse L. M., Pierson F. B., Zhang X. C., Laflen J. M., and Simanton J. R.
1995, Soil component. USDA-Water Erosion Prediction Project: Hillslope Profile and Watershed Model
Documentation. Chapter 7, pp. 7.1–7.45. Rep. No. 10. USDA-ARS National Soil Erosion Research Laboratory,
West Lafayette, IN. [Open](./resources/Alberts_1995.pdf) [Link](https://www.ars.usda.gov/midwest-area/west-lafayette-in/national-soil-erosion-research/docs/wepp/wepp-model-documentation/)

Bogomolov, A. I. and K. A. Mikhaylov (1972). Hydravlica. Moscow, Stroyizdat.

Maerker, M. (2001). Regionale Erosionsmodellierung unter Verwendung des Konzepts der Erosion Response Units (ERU) am Beispiel zweier Flusseinzugsgebiete im südlichen Afrika. Chemisch-Geowissenschaftlichen Fakultät. Jena, Friedrich-Schiller-Universität Jena. Dr. rer. nat.: 212.

Smerdon, E. T., & Beasley, R. P. (1961). Critical tractive forces in cohesive soils. Agricultural Engineering, 42(1), 26-29.

Sidorchuk, A. (1999). "Dynamic and static models of gully erosion." CATENA 37(3): 401-414.